---
title: "Kramer_etal_Liver_ILCs"
author: "Dillon Corvino"
date: "17/02/2025"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    highlight: tango
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

Built with R version `r getRversion()`

## Setup {.tabset}

### Dataset information
```{r Dataset_Info}


```

### Environment
```{r setup}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,  
  eval = TRUE, 
  tidy = TRUE
)

knitr::opts_knit$set(
  root.dir = "../"
)


# Environment Set up
rm(list = ls()) #Clean workspace
cat("\014")     #Clean Console
gc() # Free memory

# pipeline variables
long.compute <- FALSE
do.plot <- TRUE

# Establish working directory
library("rstudioapi")

# Set working directory to source file location
setwd(dirname(getActiveDocumentContext()$path))
setwd("..")

# Create output directories & load custom functions & colour scheme
source("scripts/Setup.R", local = knitr::knit_global())

# load some packages explicitly 
library(Seurat)
library(SeuratExtend)
library(scCustomize)
library(dplyr)
library(tibble)


```

### Load dataset & filter dataset
```{r load_dataset}

if(long.compute){
  
  
  # read in raw data
  raw_counts <- read.csv("data/raw_counts.csv")
  
  # make geneID the rownames
  raw_counts <- raw_counts %>% 
    tibble::column_to_rownames(var = "X") 
  
  # make cell ID match across metadata and counts tables
  colnames(raw_counts) <- gsub("\\.", "-", colnames(raw_counts))
  
  
  
  metadata <- read.csv("data/metadata.csv")
  head(metadata)
  raw_counts[1:10, 1:10]  

  
  
  
  
  seurat.object <- Seurat::CreateSeuratObject(counts = raw_counts, 
                                              meta.data = metadata)
  
  seurat.object #21,144 genes across 5,468 cells
  head(seurat.object@meta.data)
  
  table(seurat.object@meta.data$celltype)
    table(seurat.object@meta.data$orig.ident) # sample 2 = 3,518 and sample 3 = 1,950 
    
    # save the orig.ident metadata as sample ID
    seurat.object@meta.data$SampleID <- seurat.object@meta.data$orig.ident
  
    unique(seurat.object@meta.data$SampleID)
    
    seurat.object@meta.data <- seurat.object@meta.data %>%
      dplyr::mutate(tissue = case_when(SampleID == "Sample2" ~ "LTX_Liver", 
                                       SampleID == "Sample3" ~ "Ctrl_Liver", 
                                       TRUE ~ "ERROR"))
    
    unique(seurat.object@meta.data$tissue)
  
  
  # save seurat 
    SaveH5Seurat(seurat.object, "saves/seurat_object.h5Seurat", overwrite = T)

    }else{
  
  seurat.object <- LoadH5Seurat("saves/seurat_object.h5Seurat")
  
}

```

## QC and Filter {.tabset}

### QC & filter
```{r QC_and_filter}

if(long.compute){
  
  
    # Create output dir
    output.dir <- "results/QC/"
    
    if(!dir.exists(output.dir)){
      dir.create(output.dir, 
                 recursive = T)
          }
    
    
    # set idents to sampleID
    Idents(seurat.object) <- seurat.object$tissue
    
    # set default assay 
    DefaultAssay(seurat.object) <- "RNA"
    
    # Add mitochondrial and ribosomal gene percentages
    
    seurat.object <- Add_Mito_Ribo(object = seurat.object, species = "Human", overwrite = TRUE) 
    
    # Add cell complexity/novelty QC metric
    
    seurat.object <- Add_Cell_Complexity(object = seurat.object, overwrite = TRUE)
    
    
    
    # No filtering done, just to get a view of the QC of the dataset
    
    # Setting cutoff values 
    #feature.low <- 200
    #feature.high <- 5500
    #umi.low <- 500
    #umi.high <- 45000
    #mito.high <- 25
    #ribo.high <- 2
    #complexity.high <- 0.8
    
    
    ########################
    # QC plots
    ########################
    if(do.plot){
      # nFeatures
      print(
        QC_Plots_Genes(seurat_object = seurat.object, 
                       #low_cutoff = feature.low, 
                       #high_cutoff = feature.high,
                       pt.size = 0.05) 
      )
      
      dev.copy(pdf, paste0(output.dir, "genes.pdf"))
      dev.off()
      
      
      # UMI
      print(
        QC_Plots_UMIs(seurat_object = seurat.object, 
                      #low_cutoff = umi.low, 
                      #high_cutoff = umi.high, 
                      y_axis_log = T,
                      pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, "UMI.pdf"))
      dev.off()
      
      
      # Mito percentage
      print(
        QC_Plots_Mito(seurat_object = seurat.object, 
                     # high_cutoff = mito.high,
                      pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, "mito.pdf"))
      dev.off()
      
      
      # Ribo percentage
      print(
        QC_Plots_Feature(seurat_object = seurat.object,
                         feature = "percent_ribo",
                         #high_cutoff = ribo.high,
                         pt.size = 0.05)
      )
      
      dev.copy(pdf, paste0(output.dir, "Ribo.pdf"))
      dev.off()
      
      
      # Complexity
      print(
        QC_Plots_Complexity(seurat_object = seurat.object, 
                            #high_cutoff = complexity.high,
                            pt.size = 0.05)
      ) 
      
      dev.copy(pdf, paste0(output.dir, "complexity.pdf"))
      dev.off()
      
    }
    
  
}

```


### Dim Reduction

## Normalise & integrate GEX data
```{r Normalise_and_integrate_GEX_data}

if(long.compute){
  
  # set default assay 
  DefaultAssay(seurat.object) <- "RNA"
  
  # split object up based on sequencing run / in seurat V5 these are splitting the layers
  seurat.object[["RNA"]] <- split(seurat.object[["RNA"]],  f = seurat.object$SampleID)
  
  
  
  #####################################################
  # Normalise with scTransform and integrate data
  #####################################################
  # note error with more recent version of matrixStats resulting in "useNames = NA is defunct" error. downgrade to matrixStats version 1.1.0 as a workaround
  # https://github.com/satijalab/seurat/issues/7501#issuecomment-1854571904
  #remotes::install_version("matrixStats", version="1.1.0") # restart your session and run previous scripts  
  
  
  # note RNA assay is split into run level layers, below script will apply normalisation etc layer-wise
  
  # Normalise and scale with SCTransform v2
  seurat.object <- SCTransform(seurat.object, 
                           method = "glmGamPoi", 
                           vars.to.regress = "percent_mito", 
                           vst.flavor = "v2",
                           verbose = TRUE) %>%
    RunPCA(reduction.name = "pca_sct") %>%
    RunUMAP(reduction = "pca_sct",
            dims = 1:30, 
            reduction.name = "umap.unintegrated.sct")
  
  DefaultAssay(seurat.object) <- "SCT"
  
  # integrate layers using cca integration
  seurat.object <- IntegrateLayers(object = seurat.object,
                               orig.reduction = "pca_sct",
                               new.reduction = "integrated.cca.sct",
                               method = CCAIntegration,
                               normalization.method = "SCT",
                               verbose = TRUE)
  
  # find clusters and dim reduction using integrated data
  seurat.object <- FindNeighbors(seurat.object,
                             dims = 1:30,
                             reduction = "integrated.cca.sct") %>%
    FindClusters(resolution = 1.5, 
                 cluster.name = "clusters_cca_sct") %>%
    RunUMAP(reduction = "integrated.cca.sct",
            dims = 1:30, 
            reduction.name = "umap.cca.sct")
  
}

```


## Visualise integration 
```{r vis_integration}

if(do.plot){
  
  # create output dir
  output.dir <- paste0("results/integration/")
  
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)
    
  }
  
  
  unintegrated.reduction <- "umap.unintegrated.sct"
  integrated.reduction <- "umap.cca.sct"
  title.var <- "SCT"
  
  
  # evaluate integration and dim reduction
  
  vars.plot <- c("SampleID", "celltype")
  for(i in seq_along(vars.plot)){
    
    
    # remove "." from name
    name.var <- gsub("\\.", "_", vars.plot[i])
    
    before.int <- DimPlot(seurat.object,
                          reduction = unintegrated.reduction, 
                          label = T,
                          group.by = vars.plot[i]) + NoLegend() + ggtitle(paste0("Unintegrated ", title.var, " Normalisation"))
    
    after.int <- DimPlot(seurat.object,
                         reduction = integrated.reduction, 
                         label = T,
                         group.by = vars.plot[i]) + NoLegend() + ggtitle(paste0("CCA Integration using ", title.var, " Normalisation"))
    
    print(before.int | after.int)
    dev.copy(pdf, paste0(output.dir, "UMAP_before_after_integration", name.var, ".pdf"))
    dev.off()
    
  }
  
}

```








# calc ifn respons sig and plot this across cells 










### Figure 1D - Sojka IFN response heatmap & z-score
```{r Figure_1D_Sojka}

# output dir
output.dir <- "results/figures/IFNsig/"

if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)}



# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(rstatix)
library(ggpubr)



msigdbr::msigdbr_collections()

msigdbr::msigdbr_species()

# Human signatures
Hallmarks.genes <- msigdbr::msigdbr(species = "Homo sapiens", category = "H")


hall.sigs <- unique(Hallmarks.genes$gs_name)

# Extract interferon signature
hallmark.signature <- Hallmarks.genes %>%
  dplyr::filter(gs_name == "HALLMARK_INTERFERON_ALPHA_RESPONSE") %>%
  pull(gene_symbol)

# Set default assay
DefaultAssay(seurat.object) <- "SCT"

# Set identities based on cell type
Idents(seurat.object) <- seurat.object@meta.data$celltype


    # Create pseudobulk for visualization
    pseudo.HL <- AggregateExpression(seurat.object, 
                                     assays = "SCT",
                                     return.seurat = TRUE,
                                     group.by = "celltype")
    
    # Set idents of pseudobulk object
    Idents(pseudo.HL) <- "celltype"
    
    
    # Heatmap of top 10 DEGs per cluster
    gex.p <- DoHeatmap(pseudo.HL, 
                       features = hallmark.signature, 
                       draw.lines = FALSE, 
                       raster = FALSE)
    
    print(gex.p)
    
    # ggsave(filename = paste0(output.dir, "Heatmap_top_10_DEGs_", cluster.var.clean, "_pseudobulk.pdf"),
    #        plot = gex.p,
    #        device = "pdf",
    #        width = 6, height = 5)


    
    
    
    seurat.object <- AddModuleScore(seurat.object, features = list("IFNsig" = hallmark.signature), name = "IFNsig")
    
    head(seurat.object@meta.data)
    VlnPlot_scCustom(seurat.object, 
                     "IFNsig1", group.by = "celltype", split.by = "tissue", pt.size = 0)
    
    
    
    DimPlot_scCustom(seurat.object, reduction = "umap.cca.sct")
    
    
    FeaturePlot_scCustom(seurat.object, features = "IFNsig1", reduction = "umap.cca.sct")
    
    

table(seurat.object@meta.data$tissue, seurat.object@meta.data$celltype)
    
    
    
    




























    
    
    
    
    
    heatmap.df <- sojka.ordered.counts[rownames(sojka.ordered.counts) %in% hallmark.signature, ]



######################################
# with spleen NK sample
######################################


# calculate Z-score
z_scores <- t(scale(t(heatmap.df)))
z_scores <- colSums(z_scores)

# Convert the z-scores matrix to a data frame
z_scores_df <- as.data.frame(z_scores)
colnames(z_scores_df) <- "z_scores"

# Add a column indicating the group for each sample
z_scores_df$Group <- gsub("_\\d+$", "", rownames(z_scores_df))
z_scores_df$Group <- factor(z_scores_df$Group, levels = c("Spleen_NK", "Liver_NK", "Liver_ILC"))

# stats calcualtion 
ggpubr::compare_means(z_scores ~ Group,  data = z_scores_df)

# stats to plot
my.comparisons <- list( c("Spleen_NK", "Liver_NK"), c("Spleen_NK", "Liver_ILC"), c("Liver_NK", "Liver_ILC") )


# Plot boxplot
p1 <- ggplot(z_scores_df, aes(x = Group, y = z_scores, fill = Group)) +
  geom_boxplot() +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.5) +  # Add points with jitter for better visualization
  labs(x = "Group", y = "Z-Score", title = "Boxplot of Z-Scores by Group") +
  theme_minimal() +
  ggpubr::stat_compare_means(comparisons = my.comparisons, method = "wilcox.test") +
  ggpubr::stat_compare_means(method = "kruskal.test") # add anova pvalue 
    
print(p1)
dev.copy(pdf, paste0(output.dir, "Figure_1D_Sojka.pdf"))
dev.off()

######################################
# without spleen NK sample
######################################

heatmap.df <- sojka.ordered.counts[rownames(sojka.ordered.counts) %in% hallmark.signature, ]

# remove Spleen
colnames(heatmap.df)
heatmap.df <- heatmap.df[,!grepl("Spleen.*", colnames(heatmap.df))]

# calculate Z-score
z_scores <- t(scale(t(heatmap.df)))
z_scores <- colSums(z_scores)

# Convert the z-scores matrix to a data frame
z_scores_df <- as.data.frame(z_scores)
colnames(z_scores_df) <- "z_scores"

# Add a column indicating the group for each sample
z_scores_df$Group <- sub("_[^_]+$", "", rownames(z_scores_df))
z_scores_df$Group <- factor(z_scores_df$Group, levels = c("Liver_NK", "Liver_ILC"))



# stats calcualtion 
ggpubr::compare_means(z_scores ~ Group,  data = z_scores_df)

# stats to plot
my.comparisons <- list(c("Liver_NK", "Liver_ILC"))


# Plot boxplot
library(ggplot2)
p1 <- ggplot(z_scores_df, aes(x = Group, y = z_scores, fill = Group)) +
  geom_boxplot() +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.5) +  # Add points with jitter for better visualization
    labs(x = "Group", y = "Z-Score", title = "Boxplot of Z-Scores by Group") +
  theme_minimal() +
  ggpubr::stat_compare_means(method = "t.test") 

print(p1)

dev.copy(pdf, paste0(output.dir, "Figure_1D_Boxplot_no_spleen.pdf"))
dev.off()

 
```








## Dataset overview {.tabset}

## Imputation {.tabset}

### Imputate gene expression
```{r imputation}

if(long.compute){
  
  output.dir <- "saves/"
  if(!dir.exists(output.dir)){
    dir.create(output.dir, 
               recursive = T)}
  
  
  # check that genes of interest are in variable genes list 
  pbmc.seurat # 20,957 genes across 161,159 cells
  
  DefaultAssay(pbmc.seurat) <- "SCT"
  
  var.features <- VariableFeatures(pbmc.seurat)
  
  # are genes of interest in variable genes list 
  x <- grep(paste0(all.markers, collapse = "|"), var.features, value = T)
  print(x)
  length(x)
  length(all.markers)
  
  # yes all genes of interst are in variable features
  
  pbmc.seurat <- SeuratWrappers::RunALRA(pbmc.seurat, 
                                         genes.use = var.features)
  
  # Export dataset for easy import
  SeuratDisk::SaveH5Seurat(pbmc.seurat, "saves/pbmc_seurat_imputed.h5Seurat", overwrite = TRUE)
  
  DefaultAssay(pbmc.seurat) <- "SCT"
  
}


```

## Visualise {.tabset}

### Visualise Cytotoxic molecule expression -- imputed
```{r vis_cytotoxic_across_whole_dataset_imputed}

# set assay 
DefaultAssay(pbmc.seurat) <- "alra"

cts.genes <- grep("CTS.*", rownames(pbmc.seurat), value = T)
goi <- c(cytotoxicity.markers, cts.genes)



######################
# plot featureplots
######################

output.dir <- "results/figures/FeaturePlots_imputed/"
if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)}


for(i in seq_along(goi)){
  print(
    FeaturePlot(pbmc.seurat,
                features = goi[i],
                reduction = "wnn.umap",
                order = TRUE, 
                pt.size = 1, 
                raster = TRUE) + 
      NoLegend()
  )
  
  dev.copy(pdf, paste0(output.dir, "wnnUMAP_FeaturePlot_", goi[i], ".pdf"))
  dev.off()
  
}


######################
# plot density plots
######################

output.dir <- "results/figures/DensityPlots_imputed/"
if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)}

for(i in seq_along(goi)){
  print(
    scCustomize::Plot_Density_Custom(pbmc.seurat,
                                     features = goi[i],
                                     custom_palette = batlow.pal,
                                     reduction = "wnn.umap", 
                                     pt.size = 1) + 
      NoLegend()
  )
  
  dev.copy(pdf, paste0(output.dir, "wnnUMAP_DensityPlot_", goi[i], ".pdf"))
  dev.off()
  
}


######################
# Violin Plots
######################
output.dir <- "results/figures/Stacked_VlnPlots_imputed/"
if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)}

vars.plot <- c("celltype.l1", "celltype.l2", "celltype.l3", "custom_anno")


for(i in seq_along(vars.plot)){
  print(
    scCustomize::Stacked_VlnPlot(pbmc.seurat, 
                                 features = cytotoxicity.markers,
                                 pt.size = 0, 
                                 group.by = vars.plot[i],
                                 x_lab_rotate = TRUE)
  )
  
  name.var <- gsub("\\.", "_", vars.plot[i])
  dev.copy(pdf, paste0(output.dir, "Stacked_vlnplots_cytotoxicity_markers_", name.var, ".pdf"))
  dev.off()
}


scCustomize::Stacked_VlnPlot(pbmc.seurat, 
                             features = cts.genes,
                             pt.size = 0, 
                             group.by = "custom_anno",
                             x_lab_rotate = TRUE)

dev.copy(pdf, paste0(output.dir, "Stacked_vlnplots_CTS_genes.pdf"))
dev.off()






######################
# Heatmaps
######################

output.dir <- "results/figures/Heatmaps_imputed/"
if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)}


vars.plot <- c("celltype.l1", "celltype.l2", "celltype.l3", "custom_anno")

for(i in seq_along(vars.plot)){
  Idents(pbmc.seurat) <- pbmc.seurat@meta.data[[vars.plot[i]]]
  
  average.seurat <- AverageExpression(pbmc.seurat, return.seurat = TRUE)
  
  print(
    DoHeatmap(average.seurat, 
              features = cytotoxicity.markers,
              draw.lines = FALSE, 
              raster = FALSE) + 
      scico::scale_fill_scico(palette = "batlow", 
                              direction = 1, 
                              na.value = "white")
  )
  
  name.var <- gsub("\\.", "_", vars.plot[i])
  dev.copy(pdf, paste0(output.dir, "Heatmap_cytotoxicity_markers_", name.var, ".pdf"))
  dev.off()
  
  
  print(
    DoHeatmap(average.seurat, 
              features = cts.genes,
              draw.lines = FALSE, 
              raster = FALSE) + 
      scico::scale_fill_scico(palette = "batlow", 
                              direction = 1, 
                              na.value = "white")
  )
  
  name.var <- gsub("\\.", "_", vars.plot[i])
  dev.copy(pdf, paste0(output.dir, "Heatmap_CTS_genes_", name.var, ".pdf"))
  dev.off()
  
}

```




### Subsetted vlnplots granzymes -- imputed
```{r vlnplots_granzymes_imputed}

# set assay 
DefaultAssay(pbmc.seurat) <- "alra"

goi <- c(cytotoxicity.markers)


######################
# Violin Plots
######################
output.dir <- "results/figures/Stacked_VlnPlots_imputed_fig_1/"
if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)}


for(i in 1:length(cell.groups)){
  Idents(pbmc.seurat) <- pbmc.seurat@meta.data$custom_anno
  temp.seurat <- subset(pbmc.seurat, idents = cell.groups[[i]])
  
  p1 <- scCustomize::Stacked_VlnPlot(temp.seurat, 
                                     features = goi,
                                     pt.size = 0, 
                                     group.by = "custom_anno",
                                     x_lab_rotate = 90) 
  
  for(j in 1:length(p1)){
    p1[[j]] <- p1[[j]] + theme(
      plot.margin = margin(t = 0,
                           r = 10,
                           b = 0,
                           l = 0, unit = "cm")
    )
  }
  
  print(p1)
  
  dev.copy(pdf, paste0(output.dir, "Stacked_Vlnplots_",  names(cell.groups)[i], "group_cytotoxicity_imputed_.pdf"))
  dev.off()
}


```



### Subsetted vlnplots NKG7 vs TIA1 -- Raw values -- TIA1 not found in imputed
```{r vlnplots_NKG7_vs_TIA1_imputed}

# set assay 
DefaultAssay(pbmc.seurat) <- "SCT"

goi <- c("NKG7", "TIA1")

######################
# Violin Plots
######################
output.dir <- "results/figures/PBMC_Stacked_VlnPlots_SCT_TIA1_and_NKG7/"
if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)}


for(i in 1:length(cell.groups)){
  Idents(pbmc.seurat) <- pbmc.seurat@meta.data$custom_anno
  temp.seurat <- subset(pbmc.seurat, idents = cell.groups[[i]])
  
  p1 <- scCustomize::Stacked_VlnPlot(temp.seurat, 
                                     features = goi,
                                     pt.size = 0, 
                                     group.by = "custom_anno",
                                     x_lab_rotate = 90) 
  
  for(j in 1:length(p1)){
    p1[[j]] <- p1[[j]] + theme(
      plot.margin = margin(t = 0,
                           r = 10,
                           b = 0,
                           l = 0, unit = "cm")
    )
  }
  
  print(p1)
  
  dev.copy(pdf, paste0(output.dir, "Stacked_Vlnplots_",  names(cell.groups)[i], "group_NKG7_vs_TIA1_SCT.pdf"))
  dev.off()
}





# across all annotation
Idents(pbmc.seurat) <- pbmc.seurat@meta.data$custom_anno


scCustomize::Stacked_VlnPlot(pbmc.seurat, 
                             features = goi,
                             pt.size = 0, 
                             group.by = "custom_anno",
                             x_lab_rotate = 90) 



dev.copy(pdf, paste0(output.dir, "Stacked_Vlnplots_NKG7_vs_TIA1_SCT.pdf"))
dev.off()






```



### Subsetted vlnplots CST genes
```{r vlnplots_CST_genes}


######################
# Violin Plots
######################

output.dir <- "results/figures/Stacked_VlnPlots_CST_genes/"
if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)}


######################
# for imputed genes 
######################

# set assay 
DefaultAssay(pbmc.seurat) <- "alra"

goi <- grep("^CST.", rownames(pbmc.seurat), value = T)


for(i in 1:length(cell.groups)){
  Idents(pbmc.seurat) <- pbmc.seurat@meta.data$custom_anno
  temp.seurat <- subset(pbmc.seurat, idents = cell.groups[[i]])
  
  p1 <- scCustomize::Stacked_VlnPlot(temp.seurat, 
                                     features = goi,
                                     pt.size = 0, 
                                     group.by = "custom_anno",
                                     x_lab_rotate = 90) 
  
  for(j in 1:length(p1)){
    p1[[j]] <- p1[[j]] + theme(
      plot.margin = margin(t = 0,
                           r = 10,
                           b = 0,
                           l = 0, unit = "cm")
    )
  }
  
  print(p1)
  
  dev.copy(pdf, paste0(output.dir, "Stacked_Vlnplots_",  names(cell.groups)[i], "group_CST_genes_imputed.pdf"))
  dev.off()
}


######################
# for rna genes 
######################

# set assay 
DefaultAssay(pbmc.seurat) <- "SCT"

goi <- grep("^CST.", rownames(pbmc.seurat), value = T)


for(i in 1:length(cell.groups)){
  Idents(pbmc.seurat) <- pbmc.seurat@meta.data$custom_anno
  temp.seurat <- subset(pbmc.seurat, idents = cell.groups[[i]])
  
  p1 <- scCustomize::Stacked_VlnPlot(temp.seurat, 
                                     features = goi,
                                     pt.size = 0, 
                                     group.by = "custom_anno",
                                     x_lab_rotate = 90) 
  
  for(j in 1:length(p1)){
    p1[[j]] <- p1[[j]] + theme(
      plot.margin = margin(t = 0,
                           r = 10,
                           b = 0,
                           l = 0, unit = "cm")
    )
  }
  
  print(p1)
  
  dev.copy(pdf, paste0(output.dir, "Stacked_Vlnplots_",  names(cell.groups)[i], "group_CST_genes.pdf"))
  dev.off()
}


```

### Subsetted vlnplots CTS genes
```{r vlnplots_CTS_genes}

######################
# Violin Plots
######################

output.dir <- "results/figures/Stacked_VlnPlots_CTS_genes/"
if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)}



######################
# for imputed genes 
######################

# set assay 
DefaultAssay(pbmc.seurat) <- "alra"

goi <- grep("^CTS.", rownames(pbmc.seurat), value = T)


for(i in 1:length(cell.groups)){
  Idents(pbmc.seurat) <- pbmc.seurat@meta.data$custom_anno
  temp.seurat <- subset(pbmc.seurat, idents = cell.groups[[i]])
  
  p1 <- scCustomize::Stacked_VlnPlot(temp.seurat, 
                                     features = goi,
                                     pt.size = 0, 
                                     group.by = "custom_anno",
                                     x_lab_rotate = 90) 
  
  for(j in 1:length(p1)){
    p1[[j]] <- p1[[j]] + theme(
      plot.margin = margin(t = 0,
                           r = 10,
                           b = 0,
                           l = 0, unit = "cm")
    )
  }
  
  print(p1)
  
  dev.copy(pdf, paste0(output.dir, "Stacked_Vlnplots_",  names(cell.groups)[i], "group_CTS_genes_imputed.pdf"))
  dev.off()
}

######################
# for rna genes 
######################

# set assay 
DefaultAssay(pbmc.seurat) <- "SCT"

goi <- grep("^CTS.", rownames(pbmc.seurat), value = T)


for(i in 1:length(cell.groups)){
  Idents(pbmc.seurat) <- pbmc.seurat@meta.data$custom_anno
  temp.seurat <- subset(pbmc.seurat, idents = cell.groups[[i]])
  
  p1 <- scCustomize::Stacked_VlnPlot(temp.seurat, 
                                     features = goi,
                                     pt.size = 0, 
                                     group.by = "custom_anno",
                                     x_lab_rotate = 90) 
  
  for(j in 1:length(p1)){
    p1[[j]] <- p1[[j]] + theme(
      plot.margin = margin(t = 0,
                           r = 10,
                           b = 0,
                           l = 0, unit = "cm")
    )
  }
  
  print(p1)
  
  dev.copy(pdf, paste0(output.dir, "Stacked_Vlnplots_",  names(cell.groups)[i], "group_CTS_genes.pdf"))
  dev.off()
}
```

### Heatmap CST, CTS, and Cyto genes
```{r Heatmap_CST_CTS_cyto_genes}

######################
# Heatmap plot
######################

output.dir <- "results/figures/Heatmap_CST_CTS_cytogenes/"
if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)}

########################
# For imputed genes
########################

# set assay 
DefaultAssay(pbmc.seurat) <- "alra"

Idents(pbmc.seurat) <- pbmc.seurat@meta.data$custom_anno
temp.seurat <- AverageExpression(pbmc.seurat, return.seurat = TRUE)


goi.1 <- grep("^CTS.", rownames(temp.seurat), value = T)
goi.2 <- grep("^CST.", rownames(temp.seurat), value = T)
goi <- c(cytotoxicity.markers, goi.1, goi.2)



DoHeatmap(temp.seurat, 
          features = goi,
          draw.lines = FALSE, 
          raster = FALSE) + 
  scico::scale_fill_scico(palette = "batlow", 
                          direction = 1, 
                          na.value = "white")

dev.copy(pdf, paste0(output.dir, "Heatmap_CST_CTS_cytogenes_imputed.pdf"))
dev.off()

########################
# For RNA genes
########################

# set assay 
DefaultAssay(pbmc.seurat) <- "SCT"

Idents(pbmc.seurat) <- pbmc.seurat@meta.data$custom_anno
temp.seurat <- AverageExpression(pbmc.seurat, return.seurat = TRUE)

goi.1 <- grep("^CTS.", rownames(temp.seurat), value = T)
goi.2 <- grep("^CST.", rownames(temp.seurat), value = T)
goi <- c(cytotoxicity.markers, goi.1, goi.2)


DoHeatmap(temp.seurat, 
          features = goi,
          draw.lines = FALSE, 
          raster = FALSE) + 
  scico::scale_fill_scico(palette = "batlow", 
                          direction = 1, 
                          na.value = "white")

dev.copy(pdf, paste0(output.dir, "Heatmap_CST_CTS_cytogenes.pdf"))
dev.off()

```

## Cytotoxicity module score & Correlation analysis {.tabset}

### Module score for cytotoxicity & Density plot
```{r cytotoxicity_and_density_plot}

DefaultAssay(pbmc.seurat) <- "SCT"

pbmc.seurat <- AddModuleScore(pbmc.seurat,
                              features = list(cytotoxicity.markers),
                              name = "Cytotoxicity_all")


######################
# plot density plots
######################

output.dir <- "results/figures/Fig_1/"
if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)}

scCustomize::Plot_Density_Custom(pbmc.seurat,
                                 features = "Cytotoxicity_all1",
                                 custom_palette = rev(batlow.pal),
                                 reduction = "wnn.umap", 
                                 pt.size = 1) + 
  NoLegend()

dev.copy(pdf, paste0(output.dir, "wnnUMAP_DensityPlot_cytotoxicity_allmarkers.pdf"))
dev.off()

VlnPlot_scCustom(pbmc.seurat,
                 features = "Cytotoxicity_all1",
                 pt.size = 0,
                 group.by = "custom_anno",
                 sort = "increasing",
                 log = F) + NoLegend()

dev.copy(pdf, paste0(output.dir, "VlnPlot_cytotoxicity_allmarkers_ordered.pdf"))
dev.off()

```


### Test for bimodality
```{r test_bimodality}

#BiocManager::install("diptest")
#BiocManager::install("mclust")

# Load required packages
library(Seurat)
library(diptest)
library(dplyr)



output.dir <- "results/figures/bimodality_test/"
if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)}


# Load required packages
library(Seurat)
library(diptest)
library(dplyr)
library(ggplot2)

# Extract metadata from Seurat object
metadata <- pbmc.seurat@meta.data

# Perform dip test for each unique group in 'custom_anno'
dip_results <- metadata %>%
  group_by(custom_anno) %>%
  summarise(
    Dip_Statistic = dip.test(Cytotoxicity_all1)$statistic,
    P_Value = dip.test(Cytotoxicity_all1)$p.value
  ) %>%
  ungroup()

# Apply FDR correction for multiple testing
dip_results <- dip_results %>%
  mutate(
    Adjusted_P_Value = p.adjust(P_Value, method = "fdr"),
    P_Value = ifelse(P_Value < 2.2e-16, "< 2.2e-16", formatC(P_Value, format = "e", digits = 10)), #keep precision and set to r float point limit if less than 2.2e-16
    Adjusted_P_Value = ifelse(Adjusted_P_Value < 2.2e-16, "< 2.2e-16", formatC(Adjusted_P_Value, format = "e", digits = 10)),
    Significance = ifelse(as.numeric(gsub("< ", "", Adjusted_P_Value)) < 0.05, "Significant", "Not Significant")
  )

# Print results with full precision
print(dip_results)

# Save results to a CSV file
write.csv(dip_results, paste0(output.dir, "bimodality_test_results.csv"), row.names = FALSE, quote = TRUE)

# Create density plots for each subset
for (subset in unique(metadata$custom_anno)) {
  subset_data <- metadata %>% filter(custom_anno == subset) %>% pull(Cytotoxicity_all1)
  
  # Skip plotting if the subset has fewer than 5 values (density estimation requires more points)
  if (length(subset_data) < 5) next 
  
  # Create and save the histogram with density plot
  p <- ggplot(data.frame(Cytotoxicity_all1 = subset_data), aes(x = Cytotoxicity_all1)) +
    geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "blue", alpha = 0.5) +
    geom_density(color = "red", lwd = 1.5) +
    ggtitle(paste("Histogram of Cytotoxicity_all1 (", subset, ")", sep = "")) +
    xlab("Cytotoxicity Score") +
    ylab("Density") +
    theme_minimal()
  
  print(p)  # Display plot
  
  # Save the plot as an image
  ggsave(filename = paste0(output.dir, "density_plot_", subset, ".png"), plot = p, width = 6, height = 4)
}


```




### Correlation scatters
```{r correlation_scatter}

output.dir <- "results/figures/Correlation_scatter/"
if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)}


DefaultAssay(pbmc.seurat) <- "SCT"
Idents(pbmc.seurat) <- pbmc.seurat@meta.data$custom_anno

avg.seurat <- AverageExpression(pbmc.seurat, 
                                group.by = "custom_anno",
                                return.seurat = TRUE)

# really unsure why the annotation gets messed up. but use below to reset the annotation to custom_anno grouping
avg.seurat@meta.data$orig.ident <- rownames(avg.seurat@meta.data)
Idents(avg.seurat) <- avg.seurat@meta.data$orig.ident



goi <- list()

for(i in seq_along(cytotoxicity.markers)){
  goi[[i]] <- cytotoxicity.markers[-i]
  name.var <- paste0("cytotoxicity_minus_", cytotoxicity.markers[i])
  names(goi)[i] <- name.var
}

goi[[length(goi)+1]] <- cytotoxicity.markers
names(goi)[length(goi)] <- "Full_cytotoxicity_sig"



avg.seurat <- AddModuleScore(avg.seurat,
                             features = goi, 
                             name = names(goi))


end.point <- length(colnames(avg.seurat@meta.data))
start.point <- end.point-length(goi)
start.point <- start.point + 1
colnames(avg.seurat@meta.data)[start.point:end.point]


library(stringr)
string.var <- colnames(avg.seurat@meta.data)[start.point:end.point]
new_string <- str_remove(string.var, ".$")
colnames(avg.seurat@meta.data)[start.point:end.point] <- new_string


for(i in seq_along(goi)){
  if(names(goi)[i] == "Full_cytotoxicity_sig"){next}
  
  feature.1.var <- names(goi)[i]
  feature.2.var <- gsub("cytotoxicity_minus_", "", feature.1.var)
  
  print(
    FeatureScatter(avg.seurat, 
                   feature1 = feature.1.var,
                   feature2 = feature.2.var,
                   pt.size = 5,
                   span = TRUE) + scale_x_continuous(limits = c(-1, 1.5))
  )
  
  dev.copy(pdf, paste0(output.dir, "Cor_scatter_", feature.1.var, "_vs_", feature.2.var, ".pdf"))
  dev.off()
}



# cor score without pDC for GZMB sig 
Idents(avg.seurat)
temp.seurat <- subset(avg.seurat, idents = "pDC", invert = T)

FeatureScatter(temp.seurat, 
               feature1 = "cytotoxicity_minus_GZMB",
               feature2 = "GZMB",
               pt.size = 5,
               span = TRUE) + scale_x_continuous(limits = c(-1, 1.5))

dev.copy(pdf, paste0(output.dir, "Cor_scatter_Cytotoxicity_minus_GZMB_vs_GZMB_without_pDC_subset.pdf"))
dev.off()



```


### Correlation Granzymes vs Granzymes
```{r Correlation_granzymes_vs_granzymes}

# Create results directories
output.dir <- "results/figures/Correlation_Granzyme_vs_Granzyme/"

if(!dir.exists(output.dir)){
  dir.create(output.dir,
             recursive = TRUE)}

# load package
require(corrplot)

##########################
# Calculate correlation
##########################


# Function
cor.mtest <- function(mat, ...) {
  n <- ncol(mat)
  p.mat <- matrix(NA, n, n)
  diag(p.mat) <- 0
  for (i in 1:(n - 1)) {
    for (j in (i + 1):n) {
      tmp <- cor.test(mat[, i], mat[, j], ...)
      p.mat[i, j] <- p.mat[j, i] <- tmp$p.value
    }
  }
  colnames(p.mat) <- rownames(p.mat) <- colnames(mat)
  p.mat
}


########################
# Normalised RNA values
########################

DefaultAssay(pbmc.seurat) <- "SCT"
Idents(pbmc.seurat) <- pbmc.seurat@meta.data$celltype.l1
table(pbmc.seurat@meta.data$custom_anno, pbmc.seurat@meta.data$celltype.l1)

# Perform analysis as all cytotoxic cells (i.e CD4/CD8/NK), then CD4, CD8, or NK individual correlation plots
group.list <- list(cytotoxic_cells = c("CD8_T", "NK"), 
                   CD8_cells = c("CD8_T"), 
                   NK_cells = c("NK"))


for(i in 1:length(group.list)){
  
  # subset dataset for particular cell types
  group.name <- names(group.list)[i]
  temp.seurat <- subset(pbmc.seurat, idents = group.list[[i]])
  
  
  # extract RNA info
  RNA.data <- temp.seurat@assays$SCT@data
  
  # reduce dataset to just cytotoxic genes
  index.vec <- grep(paste0("^", cytotoxicity.markers, "$", collapse = "|"), rownames(RNA.data))
  
  RNA.data <- RNA.data[index.vec, ]
  RNA.data <- t(as.matrix(RNA.data))
  head(RNA.data)
  rownames(RNA.data)
  
  # Calculate correlation
  corr.val <- cor(RNA.data)
  
  # Plot correlation
  
  print(
    corrplot(corr.val,
             method = "color", 
             type = "full", 
             order = "hclust", 
             tl.col = "black", 
             diag = F))
  
  dev.copy(pdf,
           paste0(output.dir, "Corplot_RNA_", group.name, ".pdf"))
  dev.off()
  
  
  # matrix of the p-value of the correlation
  p.mat <- cor.mtest(RNA.data)
  
  print(
    corrplot(corr.val,
             method = "color", 
             type = "full", 
             order = "hclust", 
             tl.col = "black", 
             diag = FALSE,
             p.mat = p.mat,
             sig.level = 0.01, 
             insig = "blank")
  )
  
  dev.copy(pdf,
           paste0(output.dir, "Corplot_RNA_", group.name, "_sig_values_only.pdf"))
  dev.off()
  
}


```

### Correlation CTS and CST genes vs Granzymes
```{r Correlation_CTS_CST_vs_Granzymes}

# Create results directories
output.dir <- "results/figures/Correlation_CTS_CST_vs_Granzymes/"

if(!dir.exists(output.dir)){
  dir.create(output.dir,
             recursive = TRUE)}

# load package
require(corrplot)

##########################
# Calculate correlation
##########################


# Function
cor.mtest <- function(mat, ...) {
  n <- ncol(mat)
  p.mat <- matrix(NA, n, n)
  diag(p.mat) <- 0
  for (i in 1:(n - 1)) {
    for (j in (i + 1):n) {
      tmp <- cor.test(mat[, i], mat[, j], ...)
      p.mat[i, j] <- p.mat[j, i] <- tmp$p.value
    }
  }
  colnames(p.mat) <- rownames(p.mat) <- colnames(mat)
  p.mat
}


########################
# Normalised RNA values
########################

DefaultAssay(pbmc.seurat) <- "SCT"
Idents(pbmc.seurat) <- pbmc.seurat@meta.data$celltype.l1
table(pbmc.seurat@meta.data$custom_anno, pbmc.seurat@meta.data$celltype.l1)

# Perform analysis as all cytotoxic cells (i.e CD8/NK), then CD4, CD8, or NK individual correlation plots
group.list <- list(cytotoxic_cells = c("CD8_T", "NK"), 
                   CD8_cells = c("CD8_T"), 
                   NK_cells = c("NK"))


##################################################
# Analysis for CTS genes vs granzyme genes 
##################################################

cts.genes <- grep("^CTS.", rownames(pbmc.seurat), value = T)
goi <- c(cytotoxicity.markers, cts.genes)


for(i in 1:length(group.list)){
  
  # subset dataset for particular cell types
  group.name <- names(group.list)[i]
  temp.seurat <- subset(pbmc.seurat, idents = group.list[[i]])
  
  
  # extract RNA info
  RNA.data <- temp.seurat@assays$SCT@data
  
  # reduce dataset to just cytotoxic genes
  index.vec <- grep(paste0("^", goi, "$", collapse = "|"), rownames(RNA.data))
  
  RNA.data <- RNA.data[index.vec, ]
  RNA.data <- t(as.matrix(RNA.data))
  head(RNA.data)
  rownames(RNA.data)
  
  # Calculate correlation
  corr.val <- cor(RNA.data)
  
  # Plot correlation
  
  print(
    corrplot(corr.val,
             method = "color", 
             type = "full", 
             order = "hclust", 
             tl.col = "black", 
             diag = F))
  
  dev.copy(pdf,
           paste0(output.dir, "Corplot_RNA_", group.name, "_CTS_vs_Granzyme.pdf"))
  dev.off()
  
  
  # matrix of the p-value of the correlation
  p.mat <- cor.mtest(RNA.data)
  
  print(
    corrplot(corr.val,
             method = "color", 
             type = "full", 
             order = "hclust", 
             tl.col = "black", 
             diag = FALSE,
             p.mat = p.mat,
             sig.level = 0.01, 
             insig = "blank")
  )
  
  
  dev.copy(pdf,
           paste0(output.dir, "Corplot_RNA_", group.name, "_CTS_vs_Granzyme_sig_values_only.pdf"))
  dev.off()
  
  
}



##################################################
# Analysis for CST genes vs granzyme genes 
##################################################
cst.genes <- grep("^CST.", rownames(pbmc.seurat), value = T)
goi <- c(cytotoxicity.markers, cst.genes)

for(i in 1:length(group.list)){
  
  # subset dataset for particular cell types
  group.name <- names(group.list)[i]
  temp.seurat <- subset(pbmc.seurat, idents = group.list[[i]])
  
  
  # extract RNA info
  RNA.data <- temp.seurat@assays$SCT@data
  
  # reduce dataset to just cytotoxic genes
  index.vec <- grep(paste0("^", goi, "$", collapse = "|"), rownames(RNA.data))
  
  RNA.data <- RNA.data[index.vec, ]
  RNA.data <- t(as.matrix(RNA.data))
  head(RNA.data)
  rownames(RNA.data)
  
  # Calculate correlation
  corr.val <- cor(RNA.data)
  
  # Plot correlation
  
  print(
    corrplot(corr.val,
             method = "color", 
             type = "full", 
             order = "hclust", 
             tl.col = "black", 
             diag = F))
  
  dev.copy(pdf,
           paste0(output.dir, "Corplot_RNA_", group.name, "_CST_vs_Granzyme.pdf"))
  dev.off()
  
  
  # matrix of the p-value of the correlation
  p.mat <- cor.mtest(RNA.data)
  
  print(
    corrplot(corr.val,
             method = "color", 
             type = "full", 
             order = "hclust", 
             tl.col = "black", 
             diag = FALSE,
             p.mat = p.mat,
             sig.level = 0.01, 
             insig = "blank")
  )
  
  
  dev.copy(pdf,
           paste0(output.dir, "Corplot_RNA_", group.name, "_CST_vs_Granzyme_sig_values_only.pdf"))
  dev.off()
  
  
}


```
 
 
 
## Co-expression boolean {.tabset}



### Co-expression boolean upset plots
```{r boolean_coexpression_upset_plots}

# note use ComplexUpset not UpsetR package
#install.packages('ComplexUpset')
library(ComplexUpset)

# Create results directories
output.dir <- "results/figures/PBMC_dataset_boolean_upset_plots/"

if(!dir.exists(output.dir)){
  dir.create(output.dir,
             recursive = TRUE)}

################################################
# Boolean expression patterns within cell types
################################################

# determine thresholds for expression 
DefaultAssay(pbmc.seurat) <- "SCT"

theshold.val <- 1

for(i in seq_along(cytotoxicity.markers)){
  print(
  scCustomize::VlnPlot_scCustom(pbmc.seurat, 
                              feature = cytotoxicity.markers[i], 
                              pt.size = 0, 
                              group.by = "custom_anno") + 
  NoLegend() +
  geom_hline(yintercept = theshold.val, 
             linetype = "dashed",
             color = "red")
  )
  dev.copy(pdf, paste0(output.dir, "Threshold_val_determination_", cytotoxicity.markers[i], ".pdf"))
  dev.off()
  
}



########################################
# perform analysis for CD8 T cells
########################################

# CD4 CTL
# CD8 EM
# NK Bright
# NK Dim
#gd 
#gdVD2
#MAIT


# Determine expression status
DefaultAssay(pbmc.seurat) <- "SCT"
Idents(pbmc.seurat) <- pbmc.seurat@meta.data$custom_anno
unique(pbmc.seurat@meta.data$custom_anno)


# populations to loop over
subset.var <- c("CD8_Naive", "CD8_CM", "dnT", "CD8_EM", "CD4_CTL", "NK_Bright", "NK_Dim", "gdT_V9D2", "gdT", "MAIT")

for(j in seq_along(subset.var)){
  
  print(paste0("Performing analysis for ", subset.var[j]))
  
  # subset dataset 
  temp.seurat <- subset(pbmc.seurat, idents = subset.var[j])
  
  # quantify granzyme expression
  for(i in seq_along(cytotoxicity.markers)){
    logic.vec <- rownames(temp.seurat@assays$SCT@data) == cytotoxicity.markers[i]
    goi.expression <- temp.seurat@assays$SCT@data[logic.vec, ]
    expression.logic <- goi.expression > theshold.val
    sum(names(expression.logic) != rownames(temp.seurat@meta.data)) # same order, nice and easy 
    
    temp.seurat@meta.data[[paste0(cytotoxicity.markers[i], "_status")]] <- expression.logic
    
  }
  
  # extract and format metadata
  meta.data <- temp.seurat@meta.data %>%
    dplyr::select(ends_with("_status"))
  
  input_data <- data.frame(lapply(meta.data, as.integer))
  
  rownames(input_data) <- rownames(meta.data)
  
  colnames(input_data) <- gsub("_status", "", colnames(input_data))
  
  print(
    # plot upset plot 
    ComplexUpset::upset(data = input_data, 
                        n_intersections = 10,
                        intersect = rev(c("GZMA", "GZMB", "GZMH", "GZMK", "GZMM", "GNLY", "PRF1", "NKG7")),
                        name = "cytotoxic profile",
                        width_ratio = 0.15,
                        height_ratio = 0.5,
                        min_size = 1,
                        keep_empty_groups = TRUE,
                        sort_intersections = 'descending',
                        sort_sets = FALSE,
                        base_annotations = list('Intersection size' = intersection_size(
                          text_mapping = aes(label = paste0(round((!!get_size_mode("exclusive_intersection")/nrow(input_data))*100, 2), "%",
                                                            "\n",
                                                            "(",
                                                            !!get_size_mode("exclusive_intersection"),
                                                            ")")))
                          
                          + annotate(
                            geom = 'text',
                            x = Inf, 
                            y = Inf,
                            label = paste('Total:', nrow(input_data)),
                            vjust = 1,
                            hjust = 1)
                          
                          + ylab('Intersection size')
                        ), 
                        set_sizes = (upset_set_size()
                                     + theme(axis.ticks.x = element_line())
                                     + scale_y_continuous(trans = reverse_log_trans(),
                                                          labels = log10)
                                     + ylab("log10(set size)"))
    )
    
  )
  
  
  dev.copy(pdf, paste0(output.dir, "Upset_plot_", subset.var[j], "_top10_intersects.pdf"))
  dev.off()
  
  
}





```
 
 
### Co-expression boolean heatmap plots

```{r boolean_heatmap_plots}



# Create results directories
output.dir <- "results/figures/PBMC_dataset_boolean_heatmaps/"

if(!dir.exists(output.dir)){
  dir.create(output.dir,
             recursive = TRUE)}




library(gplots)

######################################################################
# heatmaps of non-scaled values showing co-expression patterns 
######################################################################

# set up object
DefaultAssay(pbmc.seurat) <- "SCT"
Idents(pbmc.seurat) <- pbmc.seurat@meta.data$custom_anno


# populations to loop over
subset.var <- c("CD8_Naive", "CD8_CM", "dnT", "CD8_EM", "CD4_CTL", "NK_Bright", "NK_Dim", "gdT_V9D2", "gdT", "MAIT")

table(pbmc.seurat@meta.data$custom_anno)

# set a custom breaks so all heatmaps are on same colour scale and can be compared
length(batlow.pal)

custom_breaks <- seq(from = 0, to = 6, length.out = 101)


for(i in seq_along(subset.var)){
  
  # subset dataset
  temp.seurat <- subset(pbmc.seurat, idents = subset.var[i])
  
  # scale dataset
  temp.seurat <- ScaleData(temp.seurat)
  
  
  # extract values 
  data_values <- FetchData(temp.seurat,
                           vars = cytotoxicity.markers, 
                           layer = "data")
  
  # remove any NA values
  data_values <- data_values %>%
    filter(!all(is.na(.)))  # Keep rows where not all values are NA
  
  
  # set random sample value, min of 2,000 or entire dataset
  n.val <- min(2000, nrow(data_values))  
  
  
  # Randomly sample rows
  sampled_df <- data_values %>% 
    sample_n(size = n.val, replace = FALSE)  # Sample without replacement
  
  # flip dataframe
  sampled_df <- t(sampled_df)
  
  # if value below threshold value, set to 0
  sampled_df[sampled_df < theshold.val] <- 0
  
  
  # Create heatmap
  print(
    heatmap.2(as.matrix(sampled_df),   # Convert data frame to matrix
              trace = "none",   # Do not display trace lines
              col = batlow.pal,  
              breaks = custom_breaks,
              scale = "none",   # Do not scale rows or columns
              dendrogram = "column",  
              Rowv = FALSE,     
              Colv = TRUE,     
              margins = c(10, 10),  # Adjust margins for row and column labels
              labRow = rownames(sampled_df),
              labCol = TRUE,
              na.rm=TRUE,
              na.color = "white",
              xlab = "Cells",
              ylab = "Cytotoxicity Markers",
              main = paste0(subset.var[i], " not scaled"),
              distfun = function(x){dist(x , method = "euclidean")},
              hclustfun = function(x){hclust(x , method = "ward.D2")}
    )
  )
  
  dev.copy(pdf, paste0(output.dir, "Heatmap_not_scaled_", subset.var[i], "_only.pdf"))
  dev.off()
  
  
}




```
 
 
 
 ### TEMP KEEP< BUT PROBABLY REMOVE this chunk
 
### Visualise NKG7 and TIA1 expression in bulk RNAseq datasets
```{r bulk_RNAseq_datasets_NKG7_vs_TIA1}


library(ggplot2)


Monaco_data <- celldex::MonacoImmuneData()



# Extract gene expression data
gene_expression <- assay(Monaco_data)
gene_expression <- as.data.frame(gene_expression)

gene_expression$gene <- rownames(gene_expression)


grep("TIA1", rownames(gene_expression), value = T)
grep("NKG7", rownames(gene_expression), value = T)

gene_expression <- gene_expression[rownames(gene_expression) %in% c("NKG7", "TIA1"), ]


library(tidyr)
library(ggplot2)

# Convert data to long format
df_long <- pivot_longer(gene_expression, 
                        cols = -gene, 
                        names_to = "Cell_Type",
                        values_to = "Expression")


unique(df_long$Cell_Type)



# Remove alphanumeric code and first underscore from Cell_Type
df_long$Cell_Type <- gsub("^[[:alnum:]]{4}_", "", df_long$Cell_Type)

# Display unique cell types after modification
unique(df_long$Cell_Type)



# Calculate average expression value for NKG7 for each cell type
average_expression <- df_long %>%
  filter(gene == "NKG7") %>%
  group_by(Cell_Type) %>%
  summarise(Avg_Expression_NKG7 = mean(Expression))

# Reorder levels of Cell_Type factor based on average expression value for NKG7
df_long$Cell_Type <- factor(df_long$Cell_Type, levels = average_expression$Cell_Type[order(average_expression$Avg_Expression_NKG7, decreasing = T)])

# Plot boxplot
ggplot(df_long, aes(x = Cell_Type, y = Expression, fill = gene)) +
  geom_boxplot() +
  geom_jitter(position = position_jitter(width = 0.2), alpha = 0.5) +
  labs(x = "Cell Type", y = "Expression Value", fill = "Gene") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = c("NKG7" = "red", "TIA1" = "blue"))  # Customize colors if needed



```





 ## needs correcting 
### NKG7 correlation across all genes 
```{r correlation_of_NKG7_vs_all_genes}


# Create results directories
output.dir <- "results/figures/Correlation_goi_vs_all_genes/"

if(!dir.exists(output.dir)){
  dir.create(output.dir,
             recursive = TRUE)}


########################
# Normalised RNA values
########################

DefaultAssay(pbmc.seurat) <- "SCT"
Idents(pbmc.seurat) <- pbmc.seurat@meta.data$celltype.l1
table(pbmc.seurat@meta.data$custom_anno, pbmc.seurat@meta.data$celltype.l1)

table(pbmc.seurat@meta.data$celltype.l1)
# 25K CD8, 18K NK cells 


group.list <- list(cytotoxic_cells = c("CD8_T", "NK"), 
                   CD8_cells = c("CD8_T"), 
                   NK_cells = c("NK"))
i = 3
output.list <- list()

  for(i in length(group.list)){
  # subset dataset for particular cell types
  group.name <- names(group.list)[i]
  temp.seurat <- subset(pbmc.seurat, idents = group.list[[i]])
  
  
  # extract normalised data table
normed.data <- as.matrix(temp.seurat@assays$SCT@data)

# calculate correlation
output.list[[i]] <- sc.correlation(data.slot = normed.data,
                        goi = "NKG7",
                        remove.quantile = 0.3,
                        output.tables = paste0("results/tables/Correlation_analysis_", group.name, "/"))

names(output.list)[i] <- group.name

  }



top.genes <- output.list[[3]] %>%
  dplyr::filter(FDR < 0.05) %>%
  dplyr::top_n(50, Cor.val) %>%
  dplyr::arrange(desc(Cor.val)) %>%
  pull(GeneID)


bot.genes <- output.list[[3]] %>%
  dplyr::filter(FDR < 0.05) %>%
  dplyr::top_n(-50, Cor.val) %>%
  dplyr::arrange(Cor.val) %>%
  pull(GeneID)

goi <- c(top.genes, bot.genes)







temp.seurat <- subset(temp.seurat, downsample = 500)
temp.seurat <- ScaleData(temp.seurat)

DoHeatmap(temp.seurat, 
          group.by = "custom_anno", 
          features = goi)




scCustomize::VlnPlot_scCustom(temp.seurat, "CST7", pt.size = 0)

goi.1 <- grep("^CST.", rownames(temp.seurat), value = T)
goi.2 <- grep("^CTS.", rownames(temp.seurat), value = T)

Idents(pbmc.seurat) <- pbmc.seurat@meta.data$custom_anno
temp.seurat <- AverageExpression(pbmc.seurat, return.seurat = T)


DoHeatmap(temp.seurat, 
          draw.lines = F,
          features = c(goi.1, goi.2, cytotoxicity.markers), raster = F)


# Getting the frequency count using table
freq_table <- table(pbmc.seurat@meta.data$custom_anno)

# Converting to data frame
df <- data.frame(value = names(freq_table), frequency = as.numeric(freq_table))

# Reshaping to long format
df_long <- df %>%
  arrange(desc(frequency))  # Arranging by frequency in descending order

# Show the transformed data


df_long$value <- factor(df_long$value, levels = df_long$value)



ggplot(df_long, aes(x = value, y = frequency)) +
  geom_col(fill = "#112446") +
  theme_minimal() +
  #scale_y_log10() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))



corrplot(corr.val,
         method = "shade", 
         type = "full", 
         order = "hclust", 
         tl.col = "black", 
         diag = FALSE,
         p.mat = p.mat,
         sig.level = 0.01, 
         insig = "blank")


DefaultAssay(temp.seurat) <- "alra"
FeatureScatter(temp.seurat, 
               feature1 = "GZMK",
               feature2 = "GZMB",
               pt.size = 5,
               span = TRUE)








```



### WGCNA
```{r remedy001}

# run in terminal 
# create new conda environment for R
#conda create -n hdWGCNA -c conda-forge r-base r-essentials

# activate conda environment
#conda activate hdWGCNA

```




### analysed up to here


### Subsettting dataset for NK, CD8, and CD4 subsets and perform dim reduction 
```{r subsetting_dataset}


# Set up dataset
DefaultAssay(reference) <- "SCT"
Idents(reference) <- reference@meta.data$celltype.l1
unique(Idents(reference))


# Subset dataset
#CD8.seurat <- subset(reference, idents = "CD8 T")
#CD4.seurat <- subset(reference, idents = "CD4 T")
#NK.seurat <- subset(reference, idents = "NK")

# Make a list of seurat objects
#seurat.list <- c(CD8.seurat, CD4.seurat, NK.seurat)
#names(seurat.list) <- c("CD8_cells", "CD4_cells", "NK_cells")

NK.T.seurat <- subset(reference, idents = c("CD8 T", "CD4 T", "NK"))

seurat.list <- c(NK.T.seurat)
names(seurat.list) <- c("NK_T_cells")

rm(NK.T.seurat)


# remove single seurat objects
#rm(CD8.seurat, CD4.seurat, NK.seurat)




# Re-normalise
seurat.list <- lapply(X = seurat.list, 
                      FUN = SCTransform, 
                      assay = "SCT",
                      method = "glmGamPoi", 
                      verbose = TRUE) 


# Scale and Calculate PCA - RNA
for(i in seq_along(seurat.list)){
  seurat.list[[i]] <- seurat.list[[i]] %>%
    NormalizeData() %>%
    FindVariableFeatures() %>% 
    ScaleData() %>% 
    RunPCA()
}






# Normalise, Scale and Calculate PCA - ADT

for(i in seq_along(seurat.list)){
  
  DefaultAssay(seurat.list[[i]]) <- 'ADT'
  
  # we will use all ADT features for dimensional reduction
  # we set a dimensional reduction name to avoid overwriting the 
  VariableFeatures(seurat.list[[i]]) <- rownames(seurat.list[[i]][["ADT"]])
  
  seurat.list[[i]] <- seurat.list[[i]] %>%
    NormalizeData(normalization.method = 'CLR', 
                  margin = 2) %>% 
    ScaleData() %>% 
    RunPCA(reduction.name = 'apca')
}


seurat.list <- lapply(X = seurat.list, 
                      FUN = FindMultiModalNeighbors, 
                      reduction.list = list("pca", "apca"),
                      dims.list = list(1:30, 1:18), 
                      modality.weight.name = "Multi.weight") 

seurat.list <- lapply(X = seurat.list, 
                      FUN = RunSPCA, 
                      assay = 'SCT',
                      graph = 'wsnn') 


seurat.list <- lapply(X = seurat.list, 
                      FUN = RunUMAP, 
                      reduction = 'spca', 
                      dims = 1:50) 




DimPlot(seurat.list[[1]], group.by = "celltype.l1", reduction = "umap") 

DimPlot(seurat.list[[1]], group.by = "donor", reduction = "umap") 
DimPlot(seurat.list[[1]], group.by = "donor", reduction = "wnn.umap") 
DimPlot(seurat.list[[1]], group.by = "celltype.l2", reduction = "umap") 



DimPlot(object = NK.seurat, 
        reduction = "wnn.umap",
        group.by = "celltype.l1",
        label = TRUE, 
        label.size = 3, 
        repel = TRUE) + 
  NoLegend()









DimPlot(reference, group.by = "donor", reduction = "umap") 
DimPlot(reference, group.by = "celltype.l2", reduction = "umap") 




SaveH5Seurat(temp.seurat, 
             filename = "Exported_RDS_files/pbmc_multimodal_CD8_cells_only.h5seurat",
             overwrite = TRUE)

# remove data
rm(temp.seurat)



```

### work in progress, playground 

### heatmap
```{r heatmap_TIA_NKG7}

######################
# Heatmap plot
######################

output.dir <- "results/figures/NKG7_vs_TIA1/"
if(!dir.exists(output.dir)){
  dir.create(output.dir, 
             recursive = T)}

########################
# For RNA genes
########################

# set assay 
DefaultAssay(pbmc.seurat) <- "SCT"

goi <- c("TIA1", "NKG7")

# quick plot to look at data
VlnPlot_scCustom(pbmc.seurat,
                 features = goi,
                 pt.size = 0,
                 group.by = "celltype.l1",
                 sort = "increasing",
                 log = FALSE, 
                 same.y.lim = T) + NoLegend()

# select order for data to be in 
temp.seurat <- pbmc.seurat
temp.seurat@meta.data$celltype.l1 <- factor(temp.seurat@meta.data$celltype.l1, levels = c("NK", "other_T", "CD8_T", "other", "CD4_T", "B", "Mono", "DC"))

# ordered data
VlnPlot_scCustom(temp.seurat,
                 features = goi,
                 pt.size = 0,
                 group.by = "celltype.l1",
                 log = FALSE, 
                 same.y.lim = T) + NoLegend()

dev.copy(pdf, paste0(output.dir, "VlnPlot_NKG7_vs_TIA1.pdf"))
dev.off()



# data from gene expression atlas



```


